RProtoBuf: Efficient Cross-Language Data Serialization in R

### 2014-01-04 papers/rjournal/eddelbuettel-francois-stokely.Rnw
one incomplete round of comments
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/rprotobuf@700 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

\abstract{Modern data collection and analysis pipelines often involve
 a sophisticated mix of applications written in general purpose and
 specialized programming languages.  Protocol Buffers are a popular
 method of serializing structured data between applications---while remaining
 independent of programming languages or operating system.  The
 \CRANpkg{RProtoBuf} package provides a complete interface to this
 library.
 %TODO(ms) keep it less than 150 words.
}

@@

Modern data collection and analysis pipelines are increasingly being
built using collections of components to better manage software
complexity through reusability, modularity, and fault
isolation \citep{Wegiel:2010:CTT:1932682.1869479}.  
Data analysis patterns such as Split-Apply-Combine
\citep{wickham2011split} explicitly break up large problems into
manageable pieces.  These patterns are frequently employed with
different programming languages used for the different phases of data
analysis -- collection, cleaning, analysis, post-processing, and
presentation in order to take advantage of the unique combination of
performance, speed of development, and library support offered by
different environments.  Each stage of the data
analysis pipeline may involve storing intermediate results in a
file or sending them over the network.

Given these requirements, how do we safely share intermediate results
between different applications, possibly written in different
languages, and possibly running on different computers?  Programming
languages such as R, Julia, Java, and Python include built-in
serialization support, but these formats are tied to the specific
programming language in use and thus lock the user into a single
environment.  CSV files can be read and written by many applications
and so are often used for exporting tabular data.  However, CSV files
have a number of disadvantages, such as a limitation of exporting only
tabular datasets, lack of type-safety, inefficient text representation
and parsing, and ambiguities in the format involving special
characters.  JSON is another widely-supported format used mostly on
the web that removes many of these disadvantages, but it too suffers
from being too slow to parse and also does not provide strong typing
between integers and floating point.  Because the schema information
is not kept separately, multiple JSON messages of the same type
needlessly duplicate the field names with each message.
%
%
%
A number of binary formats based on JSON have been proposed that
reduce the parsing cost and improve the efficiency.  MessagePack
\citep{msgpackR} and BSON \citep{rmongodb} both have R interfaces, but
these formats lack a separate schema for the serialized data and thus
still duplicate field names with each message sent over the network or
stored in a file.  Such formats also lack support for versioning when
data storage needs evolve over time.

Once the data serialization needs of an application become complex
enough, developers typically benefit from the use of an
\emph{interface description language}, or \emph{IDL}.  IDLs like
Protocol Buffers \citep{protobuf}, Apache Thrift, and Apache Avro provide a compact
well-documented schema for cross-langauge data structures and
efficient binary interchange formats.  The schema can be used to
generate model classes for statically typed programming languages such
as C++ and Java, or can be used with reflection for dynamically typed
programming languages.  Since the schema is provided separately from
the encoded data, the data can be efficiently encoded to minimize
storage costs of the stored data when compared with simple
``schema-less'' binary interchange formats.

@@

Introductory section which may include references in parentheses
\citep{R}, or cite a reference such as \citet{R} in the text.

@@

%% TODO(de,ms)  What follows is oooooold and was lifted from the webpage
%%              Rewrite?
Protocol Buffers are a modern language-neutral, platform-neutral,
extensible mechanism for sharing and storing structured data.  They
have been widely adopted in industry with applications as varied as Sony
Playstations, Twitter, Google Search, Hadoop, and Open Street Map.  While
traditional IDLs were previously characterized by bloat and
complexity, Protocol Buffers is based on a simple list and records
model that is flexible and easy to use.  Some of the key features
provided by Protocol Buffers for data analysis include:

@@

Figure~\ref{fig:protobuf-distributed-usecase} illustrates an example
communication workflow with protocol buffers and an interactive R
session.  Common use cases include populating a request RPC protocol
buffer in R that is then serialized and sent over the network to a
remote server.  The server would then deserialize the message, act on
the request, and respond with a new protocol buffer over the network.

@@

Many sources compare data serialization formats and show protocol
buffers very favorably to the alternatives, such
as \citep{Sumaray:2012:CDS:2184751.2184810}

>>>>

\abstract{Modern data collection and analysis pipelines often involve
 a sophisticated mix of applications written in general purpose and
 specialized programming languages.  Protocol Buffers are a popular
 method of serializing structured data between applications---while remaining
 independent of programming languages or operating system.  The
 \CRANpkg{RProtoBuf} package provides a complete interface between this
 library and the R environment for statistical computing.
 %TODO(ms) keep it less than 150 words.
}

@@

Modern data collection and analysis pipelines are increasingly being
built using collections of components to better manage software
complexity through reusability, modularity, and fault
isolation \citep{Wegiel:2010:CTT:1932682.1869479}.
Data analysis patterns such as Split-Apply-Combine
\citep{wickham2011split} explicitly break up large problems into
manageable pieces.  These patterns are frequently employed with
different programming languages used for the different phases of data
analysis -- collection, cleaning, analysis, post-processing, and
presentation in order to take advantage of the unique combination of
performance, speed of development, and library support offered by
different environments.  Each stage of the data
analysis pipeline may involve storing intermediate results in a
file or sending them over the network.
% DE: Nice!

Given these requirements, how do we safely share intermediate results
between different applications, possibly written in different
languages, and possibly running on different computer system, possibly
spanning different operating systems?  Programming
languages such as R, Julia, Java, and Python include built-in
serialization support, but these formats are tied to the specific
% DE: need to define serialization?
programming language in use and thus lock the user into a single
environment.  CSV files can be read and written by many applications
and so are often used for exporting tabular data.  However, CSV files
have a number of disadvantages, such as a limitation of exporting only
tabular datasets, lack of type-safety, inefficient text representation
and parsing, and ambiguities in the format involving special
characters.  JSON is another widely-supported format used mostly on
the web that removes many of these disadvantages, but it too suffers
from being too slow to parse and also does not provide strong typing
between integers and floating point.  Because the schema information
is not kept separately, multiple JSON messages of the same type
needlessly duplicate the field names with each message.
%
%
%
A number of binary formats based on JSON have been proposed that
reduce the parsing cost and improve the efficiency.  MessagePack
\citep{msgpackR} and BSON \citep{rmongodb} both have R interfaces, but
these formats lack a separate schema for the serialized data and thus
still duplicate field names with each message sent over the network or
stored in a file.  Such formats also lack support for versioning when
data storage needs evolve over time, or when application logic and
requirement changes dictate update to the message format.
% DE: Need to talk about XML ?

Once the data serialization needs of an application become complex
enough, developers typically benefit from the use of an
\emph{interface description language}, or \emph{IDL}.  IDLs like
Protocol Buffers \citep{protobuf}, Apache Thrift, and Apache Avro provide a compact
well-documented schema for cross-langauge data structures and
efficient binary interchange formats.  The schema can be used to
generate model classes for statically-typed programming languages such
as C++ and Java, or can be used with reflection for dynamically-typed
programming languages.  Since the schema is provided separately from
the encoded data, the data can be efficiently encoded to minimize
storage costs of the stored data when compared with simple
``schema-less'' binary interchange formats.

@@

FIXME Introductory section which may include references in parentheses
\citep{R}, or cite a reference such as \citet{R} in the text.

@@

%% TODO(de,ms)  What follows is oooooold and was lifted from the webpage
%%              Rewrite?
Protocol Buffers can be described as a modern, language-neutral, platform-neutral,
extensible mechanism for sharing and storing structured data.  Since their
introduction, Protocol Buffers have been widely adopted in industry with
applications as varied as database-internal messaging (Drizzle), % DE: citation?
Sony Playstations, Twitter, Google Search, Hadoop, and Open Street Map.  While
% TODO(DE): This either needs a citation, or remove the name drop
traditional IDLs have at time been criticized for code bloat and
complexity, Protocol Buffers are based on a simple list and records
model that is compartively flexible and simple to use.

Some of the key features provided by Protocol Buffers for data analysis
include:

@@

Figure~\ref{fig:protobuf-distributed-usecase} illustrates an example
communication workflow with protocol buffers and an interactive R
session.  Common use cases include populating a request RPC protocol
buffer in R that is then serialized and sent over the network to a
remote server.  The server would then deserialize the message, act on
the request, and respond with a new protocol buffer over the network. The key
difference to, say, a request to an Rserve instance is that the remote server
may not even know the R language.

@@

Many sources compare data serialization formats and show protocol
buffers very favorably to the alternatives, such
as \citet{Sumaray:2012:CDS:2184751.2184810}
